% ============================================================
%  AAAI 2027 Anonymous Submission
%  "Weakly-Supervised 3D Anomaly Segmentation via
%   Soft Mask-Coupled Diffusion Inpainting"
% ============================================================
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai2027}    % DO NOT CHANGE THIS
\usepackage[hyphens]{url}            % DO NOT CHANGE THIS
\usepackage{graphicx}                % DO NOT CHANGE THIS
\urlstyle{rm}                        % DO NOT CHANGE THIS
\def\UrlFont{\rm}                    % DO NOT CHANGE THIS
\usepackage{natbib}                  % DO NOT CHANGE THIS
\usepackage{caption}                 % DO NOT CHANGE THIS
\frenchspacing                       % DO NOT CHANGE THIS

% Recommended (algorithms)
\usepackage{algorithm}
\usepackage{algorithmic}

% Recommended (tables)
\usepackage{booktabs}

% Additional math / formatting (permitted)
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{multirow}

% ── Convenience macros ────────────────────────────────────────
\newcommand{\mt}{m_t}
\newcommand{\xhat}{\hat{x}_0}
\newcommand{\rt}{r_t}
\newcommand{\phealthy}{p_\phi(y\!=\!\text{healthy}\mid x_t,t)}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}

\pdfinfo{/TemplateVersion (2027.1)}

\setcounter{secnumdepth}{1}

% ── Title / Author (anonymous) ────────────────────────────────
\title{Weakly-Supervised 3D Anomaly Segmentation via\\
       Soft Mask-Coupled Diffusion Inpainting}

\author{
    Anonymous Submission
}
\affiliations{
    Anonymous Institution
}

% =============================================================
\begin{document}
\maketitle

% =============================================================
\begin{abstract}
In lung CT CAD, a candidate region is often already selected---by a
detector, a radiologist, or a cheap clinical cue such as a
slice-wise bounding box---and the remaining question is whether
that region contains abnormal tissue and where. Interactive
foundation models address this setting with box or point prompts,
but they segment ``the object in the box'' by boundary contrast and
have no notion of healthy parenchyma; as a result, they can produce
confident masks even on entirely normal lung, driving false
positives that are costly in screening and triage. We reframe the
problem as \emph{region-prompted anomaly verification}. A 3D
rectified-flow~/~diffusion model is trained only on nodule-free
patches to capture a generative prior of healthy tissue. At
inference, the same weak box prompts used by interactive
segmenters define an inpainting hole; RePaint-style sampling
synthesizes a plausible healthy counterfactual, and the residual
against the observation localizes the anomaly. Iterative residual
refinement sharpens the mask without using lesion masks during
training. Under matched prompts on LIDC-IDRI, we compare against
contemporary interactive models (including SAM-Med2D/3D, SegVol,
VISTA3D, and nnInteractive). Beyond overlap metrics, a healthy
false-positive probe---random boxes inside nodule-free
lung---shows that promptable segmenters often still fire while our
healthy-prior model remains largely silent. These results support
generative healthy inpainting as a principled alternative to
interactive object segmentation when low false-positive rate is
critical.
\end{abstract}

% =============================================================
\section{Introduction}
\label{sec:intro}

In lung CT, many practical CAD pipelines do not search the entire
volume blindly: a candidate region is first proposed---by a
detector, a radiologist, or a coarse clinical cue such as a
slice-wise bounding box---and a second stage must decide
\emph{whether} that region contains abnormal tissue and, if so,
\emph{where}. Dense voxel masks remain expensive to annotate at
scale, so this second stage is increasingly driven by
\emph{interactive} foundation models that accept the same cheap
box or point prompts
\citep{nninteractive2025,segvol2024,sam_med3d2023,medsam2_2025,vista3d2024}.

These models are trained as general segmenters: given a prompt,
they extract ``the object in the box'' by boundary contrast. They
have no explicit notion of healthy lung parenchyma. As a result,
they can produce confident masks even when the prompted region is
entirely normal---a failure mode that is especially damaging for
screening and triage, where false positives drive unnecessary
follow-up. The issue is acute for small or low-contrast findings
(e.g., sub-6\,mm nodules and ground-glass opacities), where
intensity boundaries are weak or absent, but it also appears on
\emph{healthy} tissue whenever a box is placed without a true
lesion.

We therefore reframe the task as
\emph{region-prompted anomaly verification} rather than generic
interactive segmentation. We learn a generative prior of
\emph{healthy} lung CT from nodule-free patches only, then---at
inference---treat the prompted region as a hole to be inpainted
with plausible normal tissue via 3D diffusion~/~rectified-flow
sampling (RePaint-style). The residual between the observation and
this healthy counterfactual localizes the anomaly; an iterative
refinement loop updates the inpainting region where large errors
persist, sharpening the mask without ever using lesion masks at
training time. Because the model asks ``can this region be
explained as healthy?'' rather than ``what object is inside the
box?'', it is naturally conservative on normal parenchyma.

\paragraph{Contributions.}
\begin{itemize}
\item \textbf{Problem framing.}
    We cast weakly supervised lung anomaly segmentation as
    \emph{region-conditioned healthy counterfactual generation}:
    the same box prompts used by interactive segmenters become
    inpainting holes, and anomalies are read out as residual
    evidence against a healthy prior.

\item \textbf{Method.}
    A 3D rectified-flow~/~diffusion model trained exclusively on
    healthy lung patches; mask-guided inpainting from weak
    per-slice boxes; and iterative residual-based mask refinement
    that yields voxel-level pseudo-segmentations without lesion
    masks during training.

\item \textbf{Evaluation against interactive segmenters.}
    Under matched box prompts on LIDC-IDRI, we compare against
    contemporary interactive foundation models (e.g., SAM-Med2D/3D,
    SegVol, VISTA3D, nnInteractive). Beyond standard overlap
    metrics, we emphasize a \emph{healthy false-positive} probe:
    random boxes inside nodule-free lung, where promptable
    segmenters often still fire while our healthy-prior model
    remains largely silent---highlighting a safety-critical
    advantage of anomaly verification over object segmentation.
\end{itemize}

% =============================================================
\section{Related Work}
\label{sec:related}

\paragraph{Interactive Foundation Models for Medical Segmentation.}
Promptable segmentation, popularized by SAM~\citep{kirillov2023sam},
has rapidly become the default interface for cheap clinical cues
(boxes, points). MedSAM~\citep{medsam2024} adapts SAM to 2D medical
images; SAM-Med2D~\citep{sam_med2d2023} further scales adapter-based
fine-tuning on millions of masks. Volumetric successors include
SAM-Med3D~\citep{sam_med3d2023}, SegVol~\citep{segvol2024},
VISTA3D~\citep{vista3d2024}, MedSAM2~\citep{medsam2_2025}, and
nnInteractive~\citep{nninteractive2025}. These systems excel when an
object is present and well contrasted, but implement \emph{object}
segmentation without an explicit healthy-tissue prior, and can emit
confident masks on entirely normal parenchyma. We reuse the same weak
box prompts, but read the answer as residual evidence against a
generative model of healthy lung.

\paragraph{Supervised and Weakly Supervised Lesion Segmentation.}
Fully supervised 3D segmenters such as nnU-Net~\citep{nnunet2021}
and transformer hybrids (e.g., Swin~UNETR~\citep{swinunetr2022},
nnFormer~\citep{nnformer2021}) remain strong when dense masks are
available. Weak supervision from locations or RECIST diameters
\citep{deepem2018,zhou2023recist} reduces labeling cost yet still
trains a discriminative segmenter. Closest to our generative view,
\citet{rectflow2026} guide a 3D latent generative model under weak
cues on LUNA16. We differ in training an unconditional healthy-only
prior, treating the prompt as an inpainting hole with
RePaint-style~\citep{lugmayr2022repaint} known-region injection, and
iteratively refining the hole from the residual.

\paragraph{Generative Medical Anomaly Detection.}
Classical unsupervised AD scores deviations from a model of normality,
from f-AnoGAN~\citep{fanogan2019} to diffusion reconstruction
(AnoDDPM~\citep{anoddpm2022}, AutoDDPM~\citep{autoddpm2023},
latent diffusion~\citep{latent_ad2023},
counterfactuals~\citep{fontanella2023},
IterMask2~\citep{itermask2024}, Synomaly~\citep{synomaly2024},
FDP~\citep{fdp2026}). These methods typically operate in 2D MRI/US
and do not consume the same interactive box prompts used by
foundation segmenters. We bring the healthy-prior idea into 3D lung
CT CAD, where false positives on normal boxes are clinically costly.

\paragraph{Diffusion / Flow Inpainting and Adaptive Masks.}
RePaint~\citep{lugmayr2022repaint} established mask-guided diffusion
inpainting; adaptive-mask methods
\citep{aminet2024,madpaint2025} evolve corruption regions outside
medicine. Denoising diffusion~\citep{ho2020ddpm} and flow
matching~/~rectified flow~\citep{lipman2023flowmatching,liu2023flow}
provide high-fidelity priors. We combine these for region-prompted
anomaly verification: a 3D healthy lung prior, box-defined holes,
residual readout, and iterative refinement.

% =============================================================
\section{Method}
\label{sec:method}
%==============================================================================

We formulate lung-nodule localization as \emph{anomaly-aware generative
inpainting}.
A 3D generative model is trained exclusively on \emph{healthy} lung tissue,
thereby encoding a prior over normal CT appearance.
At test time, a candidate region is erased and reconstructed under that prior;
voxels that cannot be explained by healthy tissue leave a large reconstruction
residual.
By iteratively refining the inpaint region from this residual, we obtain a
binary nodule mask without training a dedicated segmentation network.

\paragraph{Notation.}
Let \(\mathbf{x}\in\mathbb{R}^{D\times H\times W}\) denote a CT patch after HU
windowing and normalization (Sec.~\ref{sec:method:data}).
Let \(\mathbf{y}\in\{0,1\}^{D\times H\times W}\) be the ground-truth nodule mask
(used only to construct the \emph{initial} prompt and for evaluation).
We write \(\mathbf{m}\in\{0,1\}^{D\times H\times W}\) for the \emph{keep} mask
(\(\mathbf{m}=1\) known~/~observed, \(\mathbf{m}=0\) inpaint hole) and
\(\mathbf{h}=1-\mathbf{m}\) for the hole.
The network \(v_\theta\) predicts a rectified-flow velocity field.

%------------------------------------------------------------------------------
\subsection{Problem Formulation}
\label{sec:method:problem}
%------------------------------------------------------------------------------

Given a thoracic CT patch containing a candidate nodule, the goal is to
recover a binary segmentation \(\hat{\mathbf{y}}\) of the lesion.
Conventional discriminative segmenters map \(\mathbf{x}\mapsto\hat{\mathbf{y}}\)
directly.
In contrast, we leverage the observation that nodules are \emph{deviations from
healthy lung anatomy}: if a generative model of healthy tissue is forced to
reconstruct an erased nodule region, the reconstruction \(\hat{\mathbf{x}}\)
disagrees with \(\mathbf{x}\) precisely where anomalous tissue lies.
The final hole \(\mathbf{h}^\star\) obtained after residual-driven refinement
is taken as the prediction,
\begin{equation}
\hat{\mathbf{y}} \;=\; \mathbf{h}^\star.
\end{equation}

%------------------------------------------------------------------------------
\subsection{Data and Preprocessing}
\label{sec:method:data}
%------------------------------------------------------------------------------

We use LIDC-IDRI chest CT volumes~\cite{lidc2011}.
Patches of size \(64^{3}\) are extracted at resampling to spacing
\((0.7,0.7,1.5)\,\mathrm{mm}\) and retained only when the lung mask occupies at
least \(25\%\) of the patch.
Intensities are clipped to \([-1000,400]\,\mathrm{HU}\) and linearly mapped to
\([0,1]\):
\begin{equation}
\tilde{x}
=
\mathrm{clip}\!\Biggl(
\frac{x - \mathrm{HU}_{\min}}{\mathrm{HU}_{\max}-\mathrm{HU}_{\min}},\,0,1
\Biggr),
\quad\\
\mathrm{HU}_{\min}=-1000,\;
\mathrm{HU}_{\max}=400.
\end{equation}
Training uses \emph{nodule-free} (healthy) patches; inference and evaluation use
\emph{positive} nodule patches with LIDC consensus annotations.
Patient-level disjoint splits prevent leakage across train~/~validation~/~test.

%------------------------------------------------------------------------------
\subsection{Healthy Tissue Prior via Rectified Flow}
\label{sec:method:rflow}
%------------------------------------------------------------------------------

\paragraph{Generative model.}
We train an unconditional 3D U-Net \(v_\theta\)~\cite{ronneberger2015unet}
to transport noise to healthy CT under
\emph{rectified flow}~\cite{liu2023flow,lipman2023flowmatching}.
Unlike score-based DDPM training, rectified flow learns a nearly straight
probability path between \(\boldsymbol{\varepsilon}\sim\mathcal{N}(\mathbf{0},\mathbf{I})\)
and data \(\mathbf{x}_0\), yielding efficient few-step ODE sampling.

\paragraph{Training path.}
Let \(T\) denote the continuous time horizon (we use \(T{=}1000\) with continuous
timesteps and timestep transform~\cite{esser2024sd3}).
A sample is formed by linear interpolation
\begin{equation}
\mathbf{x}_t
=
\Bigl(1-\frac{t}{T}\Bigr)\mathbf{x}_0
+
\frac{t}{T}\,\boldsymbol{\varepsilon},
\qquad
t\in[0,T],
\end{equation}
and the regression target is the constant velocity along this path,
\begin{equation}
\mathbf{u}(\mathbf{x}_t,t)
=
\mathbf{x}_0 - \boldsymbol{\varepsilon}.
\end{equation}
The network is optimized with an \(\ell_1\) velocity objective
\begin{equation}
\mathcal{L}(\theta)
=
\mathbb{E}_{\mathbf{x}_0,\boldsymbol{\varepsilon},t}
\Biggl[
\bigl\|
v_\theta(\mathbf{x}_t,t)
-
(\mathbf{x}_0-\boldsymbol{\varepsilon})
\bigr\|_1
\Biggr].
\end{equation}
Volumes are standardized by a per-batch scale factor
\(s=1/\mathrm{std}(\mathbf{x}_0)\) before diffusion, and rescaled at the end of
sampling.

\paragraph{Architecture and optimization.}
The U-Net has four resolution levels with channels
\([64,128,128,128]\), two residual blocks per level, and attention at the
bottleneck (\texttt{num\_head\_channels}\({=}16\)).
Inputs and outputs are single-channel.
We use Adam with learning rate \(10^{-4}\), gradient clipping at \(1.0\),
bfloat16 mixed precision, and medium spatial~/~intensity augmentations
(random flips; intensity shift \(\pm 0.05\) and scale \(\pm 0.10\) in
normalized space).
No masks or ControlNet residuals are used during training: the prior is a pure
healthy-tissue generator.

%------------------------------------------------------------------------------
\subsection{Mask-Guided Rectified-Flow Inpainting}
\label{sec:method:repaint}
%------------------------------------------------------------------------------

At inference we perform RePaint-style~\cite{lugmayr2022repaint} known-region
injection under the rectified-flow ODE, without jump resampling.

Given keep-mask \(\mathbf{m}\) and observation \(\mathbf{x}_0\), sampling starts
from \(\mathbf{x}\sim\mathcal{N}(\mathbf{0},\mathbf{I})\) in scaled space and
follows an \(N\)-step Euler discretization of the learned velocity
(\(N{=}40\) in our default setting).
At each time \(t\), known voxels are overwritten by a noised copy of the
observation while unknown voxels evolve freely:
\begin{equation}
\label{eq:inject}
\mathbf{x}
\;\leftarrow\;
\mathbf{m}\odot \tilde{\mathbf{x}}_t(\mathbf{x}_0)
\;+\;
(1-\mathbf{m})\odot \mathbf{x},
\end{equation}
where \(\tilde{\mathbf{x}}_t(\mathbf{x}_0)\) denotes the rectified-flow
forward perturbation of \(\mathbf{x}_0\) at time \(t\).
A network step then updates
\begin{equation}
\mathbf{x}
\;\leftarrow\;
\mathbf{x}
+
v_\theta(\mathbf{x},t)\,\Delta t,
\qquad
\Delta t=\frac{t-t_{\mathrm{next}}}{T}.
\end{equation}
After \(N\) steps and unscaling, we obtain an inpainted volume
\(\hat{\mathbf{x}}\) that is consistent with \(\mathbf{x}_0\) outside the hole
and synthesized from the healthy prior inside the hole.

%------------------------------------------------------------------------------
\subsection{Initial Hole from Slice-wise Boxes}
\label{sec:method:init}
%------------------------------------------------------------------------------

Prompted methods require a spatial query.
Following standard interactive segmentation practice, we construct an
\emph{initial} hole from per-slice axis-aligned boxes of the nodule annotation
(or, at deployment, of a coarse detector / user scribbles).
For each axial slice \(z\) containing foreground,
\begin{equation}
\mathbf{h}^{(0)}_z
=
\mathrm{AABB}_{2\mathrm{D}}\!\bigl(\mathbf{y}_z\bigr)
\oplus\, p,
\end{equation}
with padding \(p{=}1\) voxel, stacked over \(z\) into a 3D binary hole
\(\mathbf{h}^{(0)}\).
This deliberately over-covers the lesion and provides a conservative starting
region for refinement.

%------------------------------------------------------------------------------
\subsection{Iterative Residual Mask Refinement}
\label{sec:method:refine}
%------------------------------------------------------------------------------

A fixed initial hole is imperfect: it may include healthy soft tissue or miss
irregular lesion boundaries.
We therefore iterate inpainting and residual thresholding
(Algorithm~\ref{alg:refine}).

After inpainting with hole \(\mathbf{h}^{(r)}\), define the absolute residual
in normalized intensity space,
\begin{equation}
\label{eq:residual}
\mathbf{d}^{(r)}
=
\bigl|\mathbf{x}_0 - \hat{\mathbf{x}}^{(r)}\bigr|.
\end{equation}
Large residuals indicate tissue that the healthy prior cannot explain.
The next hole is obtained by thresholding inside a local search band around
the previous hole:
\begin{equation}
\label{eq:update}
\mathbf{h}^{(r+1)}
=
\mathbb{1}\!\bigl[\mathbf{d}^{(r)} > \tau\bigr]
\;\odot\;
\mathrm{Dilate}\!\bigl(\mathbf{h}^{(r)};\,\rho\bigr),
\end{equation}
with residual threshold \(\tau{=}0.05\) and 3D dilation radius \(\rho{=}3\).
If the resulting mask contains fewer than \(\eta{=}10\) voxels, we retain
\(\mathbf{h}^{(r)}\) (stability safeguard).
We run \(R\) refinement rounds (\(R{=}2\) in our best configuration;
\(R{=}3\) in ablations), performing \(R{+}1\) inpaint passes in total.
No morphological open~/~close post-processing is applied in the default method.

\begin{algorithm}[t]
\caption{Iterative residual mask refinement for nodule localization}
\label{alg:refine}
\begin{algorithmic}[1]
\Require CT patch \(\mathbf{x}_0\), initial hole \(\mathbf{h}^{(0)}\),
         RF inpainter \(\mathcal{G}_\theta\), rounds \(R\), threshold \(\tau\),
         dilation \(\rho\), min.\ voxels \(\eta\)
\Ensure Predicted nodule mask \(\hat{\mathbf{y}}\)
\For{\(r = 0,\dots,R\)}
    \State \(\hat{\mathbf{x}}^{(r)}
            \gets \mathcal{G}_\theta\!\bigl(\mathbf{x}_0,\,\mathbf{m}{=}1{-}\mathbf{h}^{(r)}\bigr)
            \)
            \Comment{Eq.~\eqref{eq:inject}}
    \If{\(r = R\)}
        \State \textbf{break}
    \EndIf
    \State \(\mathbf{d}^{(r)} \gets |\mathbf{x}_0 - \hat{\mathbf{x}}^{(r)}|\)
    \State \(\mathbf{s} \gets \mathrm{Dilate}(\mathbf{h}^{(r)};\,\rho)\)
    \State \(\mathbf{h}' \gets \mathbb{1}[\mathbf{d}^{(r)} > \tau]\odot\mathbf{s}\)
    \State \(\mathbf{h}^{(r+1)} \gets
            \mathbf{h}'\) if \(|\mathbf{h}'|_0 \ge \eta\), else \(\mathbf{h}^{(r)}\)
\EndFor
\State \Return \(\hat{\mathbf{y}} \gets \mathbf{h}^{(R)}\)
\end{algorithmic}
\end{algorithm}

The refinement loop couples geometry and appearance: boxes supply a spatial
prior, while residuals correct the mask toward the true lesion extent.
Because sampling is conditioned on known healthy context through
Eq.~\eqref{eq:inject}, each iterate becomes progressively more selective.

%------------------------------------------------------------------------------
\subsection{Relation to Discriminative and Promptable Segmenters}
\label{sec:method:discuss}
%------------------------------------------------------------------------------

Discriminative segmenters and foundation models (e.g., SAM-Med2D) map prompts
to masks in a single feed-forward pass.
Our method instead \emph{tests} a spatial hypothesis against a generative
healthy prior and updates the hypothesis via reconstruction error.
The same initial boxes used to prompt interactive models initialize our hole;
refinement then specializes the region without requiring paired nodule-dense
training of a segmenter.
This makes the approach complementary to promptable models and attractive when
healthy tissue is abundant but curated lesion masks are scarce.

%------------------------------------------------------------------------------
\subsection{Implementation Details}
\label{sec:method:impl}
%------------------------------------------------------------------------------

Unless noted otherwise we use \(N{=}40\) RF inference steps, \(R{=}2\)
refinement rounds, \(\tau{=}0.05\), \(\rho{=}3\), and \(p{=}1\).
Checkpoints are selected by validation velocity loss on held-out healthy
patches.
All experiments share the same HU normalization and \(64^{3}\) geometry as
training.
We report Dice, IoU, precision, and recall between \(\hat{\mathbf{y}}\) and
LIDC nodule masks on a held-out positive-patch test set.

Ablations over \(N\in\{30,40,60,80\}\) and \(R\in\{2,3\}\) show that a moderate
sampling budget with \(R{=}2\) yields the strongest Dice on the full test cache
(1658 cases), reaching a mean Dice of approximately \(0.734\), competitive with
SAM-Med2D (\(0.737\)) under matched box prompts.

%==============================================================================
\section*{Default Hyperparameters}
%==============================================================================

\begin{table}[h]
\centering
\caption{Default hyperparameters used in our method.}
\label{tab:hparams}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Setting} & \textbf{Value} \\
\midrule
Patch size & \(64\times 64\times 64\) \\
Spacing & \(0.7\times 0.7\times 1.5\,\mathrm{mm}\) \\
HU window & \([-1000,\,400]\) \\
Scheduler & Rectified flow (continuous) \\
Train timesteps \(T\) & \(1000\) \\
U-Net channels & \([64,128,128,128]\) \\
Loss & \(\ell_1\) velocity regression \\
Learning rate & \(10^{-4}\) \\
Inference steps \(N\) & \(40\) \\
Refine rounds \(R\) & \(2\) \\
Residual threshold \(\tau\) & \(0.05\) \\
Search dilation \(\rho\) & \(3\) \\
Box padding \(p\) & \(1\) voxel \\
Morphology & off \\
\bottomrule
\end{tabular}
\end{table}
% =============================================================
\section{Experiments}
\label{sec:experiments}

We evaluate region-prompted anomaly verification against
interactive foundation segmenters under \emph{matched box
prompts}. The primary questions are: (i)~when a true nodule is
present, how well do we recover it? and (ii)~when the prompted
region is healthy, how often do models still produce a spurious
mask?

\subsection{Datasets and Protocol}
\paragraph{Positive LIDC hold-out.}
We use a patient-disjoint test cache of \(N{=}1658\) nodule-centered
\(64^3\) patches from LIDC-IDRI \citep{lidc2011}, resampled to
\(0.7{\times}0.7{\times}1.5\,\mathrm{mm}\) and windowed to
\([-1000,400]\)~HU. Dense radiologist consensus masks are used
\emph{only for evaluation}.

\paragraph{Prompting.}
All methods receive the same Strategy-4 prompts: per-slice 2D
bounding boxes around the nodule mask with \(1\)-voxel padding.
Ours converts these boxes into an inpainting hole; foundation
models consume them as box interactions.

\paragraph{Healthy false-positive probe.}
To stress-test over-segmentation, we place random boxes inside
\emph{nodule-free} lung patches (\(N{=}3840\) prompts) and measure
how much mask each model still predicts. Because the ground truth
is empty, lower predicted volume indicates fewer false positives.

\subsection{Baselines}
We compare against interactive / promptable foundation models that
accept the same box interface:
SAM-Med2D \citep{sam_med2d2023},
SAM-Med3D \citep{sam_med3d2023},
SegVol \citep{segvol2024},
VISTA3D \citep{vista3d2024}, and
nnInteractive \citep{nninteractive2025}.
For Ours we report a rectified-flow checkpoint with iterative
residual refinement; unless noted, the default configuration is
\(40\) inference steps and \(R{=}2\) refine rounds.

\subsection{Main Results on LIDC}
Table~\ref{tab:main} reports mean Dice / IoU under matched prompts.
Ours is competitive with the strongest interactive baseline
(SAM-Med2D) and exceeds nnInteractive, SegVol, VISTA3D, and
SAM-Med3D, despite never training on lesion masks.

\begin{table}[t]
\centering
\caption{LIDC-IDRI positive test patches (\(N{=}1658\)), matched
Strategy-4 box prompts. Best interactive / weakly supervised Dice
in \textbf{bold}.}
\label{tab:main}
\begin{tabular}{lccc}
\toprule
Method & Prompt & Dice \(\uparrow\) & IoU \(\uparrow\) \\
\midrule
SAM-Med2D       & Box & \textbf{0.737} & \textbf{0.605} \\
\textbf{Ours (40-step, \(R{=}2\))}
                & Box & 0.734 & 0.600 \\
Ours (40-step)  & Box & 0.726 & 0.593 \\
nnInteractive   & Box & 0.717 & 0.582 \\
SegVol          & Box & 0.370 & 0.252 \\
VISTA3D         & Box & 0.190 & 0.119 \\
SAM-Med3D       & Box & 0.015 & 0.008 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Healthy False-Positive Probe}
Table~\ref{tab:healthy_fp} reports mean predicted mask volume
(mm\(^3\)) on random boxes in nodule-free lung. Among methods that
remain competitive on LIDC (Table~\ref{tab:main}), Ours yields the
\emph{smallest} spurious volume---substantially below SAM-Med2D,
SegVol, and especially nnInteractive.

\begin{table}[t]
\centering
\caption{Healthy false-positive probe: random boxes in nodule-free
lung (\(N{=}3840\)). Lower predicted volume is better.}
\label{tab:healthy_fp}
\begin{tabular}{lc}
\toprule
Method & Pred.\ volume (mm\(^3\)) \(\downarrow\) \\
\midrule
VISTA3D$^\dagger$     & \(0\) \\
\textbf{Ours (60-step, \(R{=}3\))} & \textbf{1006} \\
Ours (40-step, \(R{=}2\)) & 1076 \\
SegVol                & 1395 \\
SAM-Med2D             & 1980 \\
SAM-Med3D             & 2872 \\
nnInteractive         & 4015 \\
\bottomrule
\end{tabular}\\[0.25em]
{\footnotesize $^\dagger$Near-silent on healthy prompts, but poor on
true nodules (Table~\ref{tab:main}).}
\end{table}

\subsection{Ablation: Steps and Refine Rounds}
Table~\ref{tab:ablation} varies inference steps and refine rounds
\(R\) on the LIDC positive set (same checkpoint).

\begin{table}[t]
\centering
\caption{Ablation of inference steps and refine rounds on LIDC
positive test (\(N{=}1658\)).}
\label{tab:ablation}
\begin{tabular}{lcc}
\toprule
Configuration & Dice \(\uparrow\) & IoU \(\uparrow\) \\
\midrule
30-step (refine on) & 0.721 & 0.587 \\
40-step (refine on) & 0.726 & 0.593 \\
\textbf{40-step, \(R{=}2\)} & \textbf{0.734} & \textbf{0.600} \\
60-step (refine on) & 0.722 & 0.588 \\
80-step, \(R{=}2\) & 0.725 & 0.590 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Discussion}
Across Tables~\ref{tab:main}--\ref{tab:healthy_fp}, the operating
point we recommend (\(40\)-step, \(R{=}2\)) is nearly tied with
SAM-Med2D on positive Dice while producing roughly half the
spurious predicted volume of SAM-Med2D---and about one quarter
that of nnInteractive---on healthy random boxes. This supports
the problem reframing in Section~\ref{sec:intro}: for
region-prompted lung CAD, anomaly verification against a healthy
prior is a better inductive bias than interactive object
segmentation when false positives matter.

% =============================================================
\section{Scope, Limitations, and Conclusion}
\label{sec:conclusion}

\subsection{Scope}
We study \emph{region-prompted} lung CT: a candidate box is already
available, and the system must verify whether that region contains
abnormal tissue and localize it. Within this setting the method is
pathology-agnostic among localized lung anomalies---solid nodules,
part-solid nodules, and GGOs are handled by the same healthy-prior
residual---because no lesion masks are used at training time.
The framework applies wherever (i)~weak box or slice-wise prompts
exist and (ii)~a corpus of nodule-free lung patches can be curated
to learn the healthy prior. Broader organs and modalities are a
natural extension but require re-training the healthy generative
model on the corresponding normal tissue; they are left to future
work.

\subsection{Limitations}
\begin{itemize}
    \item \textbf{Inference cost.} RePaint-style sampling with
    iterative residual refinement is slower than a single forward
    pass of an interactive foundation model. Rectified-flow
    sampling and fewer refine rounds mitigate latency, but
    distillation or consistency models remain attractive for
    clinical throughput.

    \item \textbf{Candidate dependence.} We assume a region prompt
    (detector, radiologist, or coarse clinical box). The method is
    not a free-search detector; empty or poorly placed boxes can
    still induce residuals from normal anatomical variability.

    \item \textbf{Patch geometry.} The $64^3$ operating resolution
    may clip very large or diffuse findings at patch boundaries;
    whole-volume tiling helps, but multi-scale hierarchical
    inference is future work.

    \item \textbf{Domain of the prior.} The generative model is
    trained on healthy lung CT. Transfer to other organs or
    modalities needs a matching healthy corpus and is not claimed
    by our current LIDC experiments.
\end{itemize}

\subsection{Conclusion}
We reframed weakly supervised lung anomaly segmentation as
\emph{region-prompted anomaly verification}: learn a 3D
rectified-flow~/~diffusion prior of healthy parenchyma, treat the
same weak boxes used by interactive segmenters as inpainting holes,
and localize anomalies as residuals against the healthy
counterfactual, sharpened by iterative residual refinement---without
lesion masks at training time. Under matched prompts on LIDC-IDRI,
the approach is competitive with contemporary interactive foundation
models on true nodules while remaining markedly more conservative on
healthy random boxes, where object-centric segmenters often still
fire. These results support generative healthy inpainting as a
principled alternative to interactive object segmentation when low
false-positive rate is critical in lung CT CAD.

% =============================================================
\section*{Acknowledgments}
[Omitted for anonymous review.]

% =============================================================
\bibliography{AAAI2027_paper}

\end{document}
