%\documentclass{article}
%\usepackage{graphicx} % Required for inserting images

%\title{Hojjat-AAAI-2027}
%\author{morteza.rezanejad }
%\date{July 2026}

%\begin{document}

%\maketitle

%\section{Introduction}

%\end{document}
% ============================================================
%  AAAI 2027 Anonymous Submission
%  "Weakly-Supervised 3D Anomaly Segmentation via
%   Soft Mask-Coupled Diffusion Inpainting"
% ============================================================
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai2027}    % DO NOT CHANGE THIS
\usepackage[hyphens]{url}            % DO NOT CHANGE THIS
\usepackage{graphicx}                % DO NOT CHANGE THIS
\urlstyle{rm}                        % DO NOT CHANGE THIS
\def\UrlFont{\rm}                    % DO NOT CHANGE THIS
\usepackage{natbib}                  % DO NOT CHANGE THIS
\usepackage{caption}                 % DO NOT CHANGE THIS
\frenchspacing                       % DO NOT CHANGE THIS

% Recommended (algorithms)
\usepackage{algorithm}
\usepackage{algorithmic}

% Recommended (tables)
\usepackage{booktabs}

% Additional math / formatting (permitted)
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{multirow}

% ── Convenience macros ────────────────────────────────────────
\newcommand{\mt}{m_t}
\newcommand{\xhat}{\hat{x}_0}
\newcommand{\rt}{r_t}
\newcommand{\phealthy}{p_\phi(y\!=\!\text{healthy}\mid x_t,t)}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}

\pdfinfo{/TemplateVersion (2027.1)}

\setcounter{secnumdepth}{1}

% ── Title / Author (anonymous) ────────────────────────────────
\title{Weakly-Supervised 3D Anomaly Segmentation via\\
       Soft Mask-Coupled Diffusion Inpainting}

\author{
    Anonymous Submission
}
\affiliations{
    Anonymous Institution
}

% =============================================================
\begin{document}
\maketitle

% =============================================================
\begin{abstract}
In lung CT CAD, a candidate region is often already selected---by a detector, a radiologist, or a cheap clinical cue such as a slice-wise bounding box---and the remaining question is whether that region contains abnormal tissue and where. Interactive foundation models address this setting with box or point prompts, but they segment ``the object in the box'' by boundary contrast and have no notion of healthy parenchyma; as a result, they can produce confident masks even on entirely normal lung, driving false positives that are costly in screening and triage. We reframe the problem as \emph{region-prompted anomaly verification}. A 3D rectified-flow / diffusion model is trained only on nodule-free patches to capture a generative prior of healthy tissue. At inference, the same weak box prompts used by interactive segmenters define an inpainting hole; RePaint-style sampling synthesizes a plausible healthy counterfactual, and the residual against the observation localizes the anomaly. Iterative residual refinement sharpens the mask without using lesion masks during training. Under matched prompts on LIDC-IDRI, we compare against contemporary interactive models (including SAM-Med2D/3D, SegVol, VISTA3D, and nnInteractive). Beyond overlap metrics, a healthy false-positive probe---random boxes inside nodule-free lung---shows that promptable segmenters often still fire while our healthy-prior model remains largely silent. These results support generative healthy inpainting as a principled alternative to interactive object segmentation when low false-positive rate is critical.
\end{abstract}

% =============================================================
\section{Introduction}
\label{sec:intro}

In lung CT, many practical CAD pipelines do not search the entire
volume blindly: a candidate region is first proposed---by a
detector, a radiologist, or a coarse clinical cue such as a
slice-wise bounding box---and a second stage must decide
\emph{whether} that region contains abnormal tissue and, if so,
\emph{where}. Dense voxel masks remain expensive to annotate at
scale, so this second stage is increasingly driven by
\emph{interactive} foundation models that accept the same cheap
box or point prompts
\citep{nninteractive2025,segvol2024,sam_med3d2023,medsam2_2025,vista3d2024}.

These models are trained as general segmenters: given a prompt,
they extract ``the object in the box'' by boundary contrast. They
have no explicit notion of healthy lung parenchyma. As a result,
they can produce confident masks even when the prompted region is
entirely normal---a failure mode that is especially damaging for
screening and triage, where false positives drive unnecessary
follow-up. The issue is acute for small or low-contrast findings
(e.g., sub-6\,mm nodules and ground-glass opacities), where
intensity boundaries are weak or absent, but it also appears on
\emph{healthy} tissue whenever a box is placed without a true
lesion.

We therefore reframe the task as
\emph{region-prompted anomaly verification} rather than generic
interactive segmentation. We learn a generative prior of
\emph{healthy} lung CT from nodule-free patches only, then---at
inference---treat the prompted region as a hole to be inpainted
with plausible normal tissue via 3D diffusion / rectified-flow
sampling (RePaint-style). The residual between the observation and
this healthy counterfactual localizes the anomaly; an iterative
refinement loop updates the inpainting region where large errors
persist, sharpening the mask without ever using lesion masks at
training time. Because the model asks ``can this region be
explained as healthy?'' rather than ``what object is inside the
box?'', it is naturally conservative on normal parenchyma.

\paragraph{Contributions.}
\begin{itemize}
\item \textbf{Problem framing.}
    We cast weakly supervised lung anomaly segmentation as
    \emph{region-conditioned healthy counterfactual generation}:
    the same box prompts used by interactive segmenters become
    inpainting holes, and anomalies are read out as residual
    evidence against a healthy prior.

\item \textbf{Method.}
    A 3D rectified-flow / diffusion model trained exclusively on
    healthy lung patches; mask-guided inpainting from weak
    per-slice boxes; and iterative residual-based mask refinement
    that yields voxel-level pseudo-segmentations without lesion
    masks during training.

\item \textbf{Evaluation against interactive segmenters.}
    Under matched box prompts on LIDC-IDRI, we compare against
    contemporary interactive foundation models (e.g., SAM-Med2D/3D,
    SegVol, VISTA3D, nnInteractive). Beyond standard overlap
    metrics, we emphasize a \emph{healthy false-positive} probe:
    random boxes inside nodule-free lung, where promptable
    segmenters often still fire while our healthy-prior model
    remains largely silent---highlighting a safety-critical
    advantage of anomaly verification over object segmentation.
\end{itemize}

% =============================================================
\section{Related Work}
\label{sec:related}

\paragraph{Supervised 3D Lesion Segmentation.}
nnU-Net \citep{nnunet2021} is the de facto upper bound for
supervised CT segmentation, auto-configuring a 3D U-Net to any
dataset. Transformer-based architectures (SwinUNETR, nnFormer)
improve on LIDC-IDRI but require full mask supervision and degrade
sharply outside their training distribution, especially on
sub-6\,mm nodules due to extreme class imbalance.

\paragraph{Weakly Supervised Lesion Segmentation.}
DeepEM \citep{deepem2018} trains 3D CNNs from location labels via
EM over latent boxes. RECIST-based supervision
\citep{zhou2023recist} converts 2D diameters into 3D pseudo-
labels. Most relevantly, \citet{rectflow2026} use a 3D latent
diffusion model's generative guidance to suppress nodule signal in
reconstruction on LUNA16---the closest existing work. Key
differences: their model is not trained on healthy-only data;
their mask is not evolved inside the reverse process; they provide
no classifier guidance, regularizer, or cross-organ/modality
evaluation.

\paragraph{Reconstruction-Based Anomaly Detection.}
AnoDDPM \citep{anoddpm2022} introduced partial diffusion with
simplex noise for brain MRI, training on healthy data and
computing reconstruction error as the anomaly score
($+25.5\%$ Dice over f-AnoGAN). Extensions include latent-space
partial diffusion \citep{latent_ad2023}, Synomaly noise for
ultrasound \citep{synomaly2024}, counterfactual generation for
brain tumours \citep{fontanella2023}, IterMask2
\citep{itermask2024}, and FDP \citep{fdp2026} for brain MRI.
All are 2D or 2D-stacked, use a fixed corruption region, and do
\emph{not} evolve the mask inside the reverse process.

\paragraph{Adaptive Mask Inpainting.}
AMI-Net \citep{aminet2024} proposes an adaptive mask generator for
industrial anomaly detection (MVTec AD, BTAD). MAD-paint
\citep{madpaint2025} introduces mask-uncertainty metrics for
diffusion inpainting. Neither is 3D, medical, nor couples the mask
inside the reverse diffusion kernel.

\paragraph{Foundation Models for Medical Segmentation.}
MedSAM \citep{medsam2024} fine-tunes SAM on medical data but
remains 2D. SAM-Med3D \citep{sam_med3d2023} retrains SAM natively
in 3D. SegVol \citep{segvol2024} is a universal interactive CT
segmenter. MedSAM2 \citep{medsam2_2025} applies SAM2's video
memory to 3D volumes. nnInteractive \citep{nninteractive2025} won
the CVPR 2025 Foundation Models for Interactive 3D Biomedical
Image Segmentation Challenge. All are boundary-detection paradigms that lack an explicit
anomaly-versus-normal generative prior; they also tend to degrade
on small or low-contrast lesions.

% =============================================================
\section{Methodology}
\label{sec:method}
%==============================================================================

We formulate lung-nodule localization as \emph{anomaly-aware generative
inpainting}.
A 3D generative model is trained exclusively on \emph{healthy} lung tissue,
thereby encoding a prior over normal CT appearance.
At test time, a candidate region is erased and reconstructed under that prior;
voxels that cannot be explained by healthy tissue leave a large reconstruction
residual.
By iteratively refining the inpaint region from this residual, we obtain a
binary nodule mask without training a dedicated segmentation network.

\paragraph{Notation.}
Let \(\mathbf{x}\in\mathbb{R}^{D\times H\times W}\) denote a CT patch after HU
windowing and normalization (Sec.~\ref{sec:method:data}).
Let \(\mathbf{y}\in\{0,1\}^{D\times H\times W}\) be the ground-truth nodule mask
(used only to construct the \emph{initial} prompt and for evaluation).
We write \(\mathbf{m}\in\{0,1\}^{D\times H\times W}\) for the \emph{keep} mask
(\(\mathbf{m}=1\) known~/~observed, \(\mathbf{m}=0\) inpaint hole) and
\(\mathbf{h}=1-\mathbf{m}\) for the hole.
The network \(v_\theta\) predicts a rectified-flow velocity field.

%------------------------------------------------------------------------------
\subsection{Problem Formulation}
\label{sec:method:problem}
%------------------------------------------------------------------------------

Given a thoracic CT patch containing a candidate nodule, the goal is to
recover a binary segmentation \(\hat{\mathbf{y}}\) of the lesion.
Conventional discriminative segmenters map \(\mathbf{x}\mapsto\hat{\mathbf{y}}\)
directly.
In contrast, we leverage the observation that nodules are \emph{deviations from
healthy lung anatomy}: if a generative model of healthy tissue is forced to
reconstruct an erased nodule region, the reconstruction \(\hat{\mathbf{x}}\)
disagrees with \(\mathbf{x}\) precisely where anomalous tissue lies.
The final hole \(\mathbf{h}^\star\) obtained after residual-driven refinement
is taken as the prediction,
\begin{equation}
\hat{\mathbf{y}} \;=\; \mathbf{h}^\star.
\end{equation}

%------------------------------------------------------------------------------
\subsection{Data and Preprocessing}
\label{sec:method:data}
%------------------------------------------------------------------------------

We use LIDC-IDRI chest CT volumes~\cite{armato2011lidc}.
Patches of size \(64^{3}\) are extracted at resampling to spacing
\((0.7,0.7,1.5)\,\mathrm{mm}\) and retained only when the lung mask occupies at
least \(25\%\) of the patch.
Intensities are clipped to \([-1000,400]\,\mathrm{HU}\) and linearly mapped to
\([0,1]\):
\begin{equation}
\tilde{x}
=
\mathrm{clip}\!\Biggl(
\frac{x - \mathrm{HU}_{\min}}{\mathrm{HU}_{\max}-\mathrm{HU}_{\min}},\,0,1
\Biggr),
\quad\\
\mathrm{HU}_{\min}=-1000,\;
\mathrm{HU}_{\max}=400.
\end{equation}
Training uses \emph{nodule-free} (healthy) patches; inference and evaluation use
\emph{positive} nodule patches with LIDC consensus annotations.
Patient-level disjoint splits prevent leakage across train~/~validation~/~test.

%------------------------------------------------------------------------------
\subsection{Healthy Tissue Prior via Rectified Flow}
\label{sec:method:rflow}
%------------------------------------------------------------------------------

\paragraph{Generative model.}
We train an unconditional 3D U-Net \(v_\theta\)~\cite{ronneberger2015unet}
to transport noise to healthy CT under
\emph{rectified flow}~\cite{liu2023flow,lipman2023flowmatching}.
Unlike score-based DDPM training, rectified flow learns a nearly straight
probability path between \(\boldsymbol{\varepsilon}\sim\mathcal{N}(\mathbf{0},\mathbf{I})\)
and data \(\mathbf{x}_0\), yielding efficient few-step ODE sampling.

\paragraph{Training path.}
Let \(T\) denote the continuous time horizon (we use \(T{=}1000\) with continuous
timesteps and timestep transform~\cite{esser2024sd3}).
A sample is formed by linear interpolation
\begin{equation}
\mathbf{x}_t
=
\Bigl(1-\frac{t}{T}\Bigr)\mathbf{x}_0
+
\frac{t}{T}\,\boldsymbol{\varepsilon},
\qquad
t\in[0,T],
\end{equation}
and the regression target is the constant velocity along this path,
\begin{equation}
\mathbf{u}(\mathbf{x}_t,t)
=
\mathbf{x}_0 - \boldsymbol{\varepsilon}.
\end{equation}
The network is optimized with an \(\ell_1\) velocity objective
\begin{equation}
\mathcal{L}(\theta)
=
\mathbb{E}_{\mathbf{x}_0,\boldsymbol{\varepsilon},t}
\Biggl[
\bigl\|
v_\theta(\mathbf{x}_t,t)
-
(\mathbf{x}_0-\boldsymbol{\varepsilon})
\bigr\|_1
\Biggr].
\end{equation}
Volumes are standardized by a per-batch scale factor
\(s=1/\mathrm{std}(\mathbf{x}_0)\) before diffusion, and rescaled at the end of
sampling.

\paragraph{Architecture and optimization.}
The U-Net has four resolution levels with channels
\([64,128,128,128]\), two residual blocks per level, and attention at the
bottleneck (\texttt{num\_head\_channels}\({=}16\)).
Inputs and outputs are single-channel.
We use Adam with learning rate \(10^{-4}\), gradient clipping at \(1.0\),
bfloat16 mixed precision, and medium spatial~/~intensity augmentations
(random flips; intensity shift \(\pm 0.05\) and scale \(\pm 0.10\) in
normalized space).
No masks or ControlNet residuals are used during training: the prior is a pure
healthy-tissue generator.

%------------------------------------------------------------------------------
\subsection{Mask-Guided Rectified-Flow Inpainting}
\label{sec:method:repaint}
%------------------------------------------------------------------------------

At inference we perform RePaint-style~\cite{lugmayr2022repaint} known-region
injection under the rectified-flow ODE, without jump resampling.

Given keep-mask \(\mathbf{m}\) and observation \(\mathbf{x}_0\), sampling starts
from \(\mathbf{x}\sim\mathcal{N}(\mathbf{0},\mathbf{I})\) in scaled space and
follows an \(N\)-step Euler discretization of the learned velocity
(\(N{=}40\) in our default setting).
At each time \(t\), known voxels are overwritten by a noised copy of the
observation while unknown voxels evolve freely:
\begin{equation}
\label{eq:inject}
\mathbf{x}
\;\leftarrow\;
\mathbf{m}\odot \tilde{\mathbf{x}}_t(\mathbf{x}_0)
\;+\;
(1-\mathbf{m})\odot \mathbf{x},
\end{equation}
where \(\tilde{\mathbf{x}}_t(\mathbf{x}_0)\) denotes the rectified-flow
forward perturbation of \(\mathbf{x}_0\) at time \(t\).
A network step then updates
\begin{equation}
\mathbf{x}
\;\leftarrow\;
\mathbf{x}
+
v_\theta(\mathbf{x},t)\,\Delta t,
\qquad
\Delta t=\frac{t-t_{\mathrm{next}}}{T}.
\end{equation}
After \(N\) steps and unscaling, we obtain an inpainted volume
\(\hat{\mathbf{x}}\) that is consistent with \(\mathbf{x}_0\) outside the hole
and synthesized from the healthy prior inside the hole.

%------------------------------------------------------------------------------
\subsection{Initial Hole from Slice-wise Boxes}
\label{sec:method:init}
%------------------------------------------------------------------------------

Prompted methods require a spatial query.
Following standard interactive segmentation practice, we construct an
\emph{initial} hole from per-slice axis-aligned boxes of the nodule annotation
(or, at deployment, of a coarse detector / user scribbles).
For each axial slice \(z\) containing foreground,
\begin{equation}
\mathbf{h}^{(0)}_z
=
\mathrm{AABB}_{2\mathrm{D}}\!\bigl(\mathbf{y}_z\bigr)
\oplus\, p,
\end{equation}
with padding \(p{=}1\) voxel, stacked over \(z\) into a 3D binary hole
\(\mathbf{h}^{(0)}\).
This deliberately over-covers the lesion and provides a conservative starting
region for refinement.

%------------------------------------------------------------------------------
\subsection{Iterative Residual Mask Refinement}
\label{sec:method:refine}
%------------------------------------------------------------------------------

A fixed initial hole is imperfect: it may include healthy soft tissue or miss
irregular lesion boundaries.
We therefore iterate inpainting and residual thresholding
(Algorithm~\ref{alg:refine}).

After inpainting with hole \(\mathbf{h}^{(r)}\), define the absolute residual
in normalized intensity space,
\begin{equation}
\label{eq:residual}
\mathbf{d}^{(r)}
=
\bigl|\mathbf{x}_0 - \hat{\mathbf{x}}^{(r)}\bigr|.
\end{equation}
Large residuals indicate tissue that the healthy prior cannot explain.
The next hole is obtained by thresholding inside a local search band around
the previous hole:
\begin{equation}
\label{eq:update}
\mathbf{h}^{(r+1)}
=
\mathbb{1}\!\bigl[\mathbf{d}^{(r)} > \tau\bigr]
\;\odot\;
\mathrm{Dilate}\!\bigl(\mathbf{h}^{(r)};\,\rho\bigr),
\end{equation}
with residual threshold \(\tau{=}0.05\) and 3D dilation radius \(\rho{=}3\).
If the resulting mask contains fewer than \(\eta{=}10\) voxels, we retain
\(\mathbf{h}^{(r)}\) (stability safeguard).
We run \(R\) refinement rounds (\(R{=}2\) in our best configuration;
\(R{=}3\) in ablations), performing \(R{+}1\) inpaint passes in total.
No morphological open~/~close post-processing is applied in the default method.

\begin{algorithm}[t]
\caption{Iterative residual mask refinement for nodule localization}
\label{alg:refine}
\begin{algorithmic}[1]
\Require CT patch \(\mathbf{x}_0\), initial hole \(\mathbf{h}^{(0)}\),
         RF inpainter \(\mathcal{G}_\theta\), rounds \(R\), threshold \(\tau\),
         dilation \(\rho\), min.\ voxels \(\eta\)
\Ensure Predicted nodule mask \(\hat{\mathbf{y}}\)
\For{\(r = 0,\dots,R\)}
    \State \(\hat{\mathbf{x}}^{(r)}
            \gets \mathcal{G}_\theta\!\bigl(\mathbf{x}_0,\,\mathbf{m}{=}1{-}\mathbf{h}^{(r)}\bigr)
            \)
            \Comment{Eq.~\eqref{eq:inject}}
    \If{\(r = R\)}
        \State \textbf{break}
    \EndIf
    \State \(\mathbf{d}^{(r)} \gets |\mathbf{x}_0 - \hat{\mathbf{x}}^{(r)}|\)
    \State \(\mathbf{s} \gets \mathrm{Dilate}(\mathbf{h}^{(r)};\,\rho)\)
    \State \(\mathbf{h}' \gets \mathbb{1}[\mathbf{d}^{(r)} > \tau]\odot\mathbf{s}\)
    \State \(\mathbf{h}^{(r+1)} \gets
            \mathbf{h}'\) if \(|\mathbf{h}'|_0 \ge \eta\), else \(\mathbf{h}^{(r)}\)
\EndFor
\State \Return \(\hat{\mathbf{y}} \gets \mathbf{h}^{(R)}\)
\end{algorithmic}
\end{algorithm}

The refinement loop couples geometry and appearance: boxes supply a spatial
prior, while residuals correct the mask toward the true lesion extent.
Because sampling is conditioned on known healthy context through
Eq.~\eqref{eq:inject}, each iterate becomes progressively more selective.

%------------------------------------------------------------------------------
\subsection{Relation to Discriminative and Promptable Segmenters}
\label{sec:method:discuss}
%------------------------------------------------------------------------------

Discriminative segmenters and foundation models (e.g., SAM-Med2D) map prompts
to masks in a single feed-forward pass.
Our method instead \emph{tests} a spatial hypothesis against a generative
healthy prior and updates the hypothesis via reconstruction error.
The same initial boxes used to prompt interactive models initialize our hole;
refinement then specializes the region without requiring paired nodule-dense
training of a segmenter.
This makes the approach complementary to promptable models and attractive when
healthy tissue is abundant but curated lesion masks are scarce.

%------------------------------------------------------------------------------
\subsection{Implementation Details}
\label{sec:method:impl}
%------------------------------------------------------------------------------

Unless noted otherwise we use \(N{=}40\) RF inference steps, \(R{=}2\)
refinement rounds, \(\tau{=}0.05\), \(\rho{=}3\), and \(p{=}1\).
Checkpoints are selected by validation velocity loss on held-out healthy
patches.
All experiments share the same HU normalization and \(64^{3}\) geometry as
training.
We report Dice, IoU, precision, and recall between \(\hat{\mathbf{y}}\) and
LIDC nodule masks on a held-out positive-patch test set.

Ablations over \(N\in\{30,40,60,80\}\) and \(R\in\{2,3\}\) show that a moderate
sampling budget with \(R{=}2\) yields the strongest Dice on the full test cache
(1658 cases), reaching a mean Dice of approximately \(0.734\), competitive with
SAM-Med2D (\(0.737\)) under matched box prompts.

%==============================================================================
\section*{Default Hyperparameters}
%==============================================================================

\begin{table}[h]
\centering
\caption{Default hyperparameters used in our method.}
\label{tab:hparams}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Setting} & \textbf{Value} \\
\midrule
Patch size & \(64\times 64\times 64\) \\
Spacing & \(0.7\times 0.7\times 1.5\,\mathrm{mm}\) \\
HU window & \([-1000,\,400]\) \\
Scheduler & Rectified flow (continuous) \\
Train timesteps \(T\) & \(1000\) \\
U-Net channels & \([64,128,128,128]\) \\
Loss & \(\ell_1\) velocity regression \\
Learning rate & \(10^{-4}\) \\
Inference steps \(N\) & \(40\) \\
Refine rounds \(R\) & \(2\) \\
Residual threshold \(\tau\) & \(0.05\) \\
Search dilation \(\rho\) & \(3\) \\
Box padding \(p\) & \(1\) voxel \\
Morphology & off \\
\bottomrule
\end{tabular}
\end{table}
% =============================================================
\section{Experiments}
\label{sec:experiments}

We evaluate region-prompted anomaly verification against
interactive foundation segmenters under \emph{matched box
prompts}. The primary questions are: (i)~when a true nodule is
present, how well do we recover it? and (ii)~when the prompted
region is healthy, how often do models still produce a spurious
mask?

\subsection{Datasets and Protocol}
\paragraph{Positive LIDC hold-out.}
We use a patient-disjoint test cache of \(N{=}1658\) nodule-centered
\(64^3\) patches from LIDC-IDRI \citep{lidc2011}, resampled to
\(0.7{\times}0.7{\times}1.5\,\mathrm{mm}\) and windowed to
\([-1000,400]\)~HU. Dense radiologist consensus masks are used
\emph{only for evaluation}.

\paragraph{Prompting.}
All methods receive the same Strategy-4 prompts: per-slice 2D
bounding boxes around the nodule mask with \(1\)-voxel padding
(the interactive-segmentation analogue of a weak clinical cue).
Ours converts these boxes into an inpainting hole; foundation
models consume them as box interactions.

\paragraph{Healthy false-positive probe.}
To stress-test over-segmentation, we place random boxes inside
\emph{nodule-free} lung patches (\(N{=}3840\) prompts) and measure
how much mask each model still predicts. Because the ground truth
is empty, lower predicted volume indicates fewer false positives.

\subsection{Baselines}
We compare against interactive / promptable foundation models that
accept the same box interface:
SAM-Med2D \citep{medsam2024},
SAM-Med3D \citep{sam_med3d2023},
SegVol \citep{segvol2024},
VISTA3D \citep{vista3d2024}, and
nnInteractive \citep{nninteractive2025} (CVPR~2025 challenge
winner).
For Ours we report a rectified-flow checkpoint with iterative
residual refinement; unless noted, the default configuration is
\(40\) inference steps and \(R{=}2\) refine rounds.

\subsection{Main Results on LIDC}
Table~\ref{tab:main} reports mean Dice / IoU under matched prompts.
Ours is competitive with the strongest interactive baseline
(SAM-Med2D) and exceeds nnInteractive, SegVol, VISTA3D, and
SAM-Med3D, despite never training on lesion masks.

\begin{table}[t]
\centering
\caption{LIDC-IDRI positive test patches (\(N{=}1658\)), matched
Strategy-4 box prompts. Best interactive / weakly supervised Dice
in \textbf{bold}.}
\label{tab:main}
\begin{tabular}{lccc}
\toprule
Method & Prompt & Dice \(\uparrow\) & IoU \(\uparrow\) \\
\midrule
SAM-Med2D       & Box & \textbf{0.737} & \textbf{0.605} \\
\textbf{Ours (40-step, \(R{=}2\))}
                & Box & 0.734 & 0.600 \\
Ours (40-step)  & Box & 0.726 & 0.593 \\
nnInteractive   & Box & 0.717 & 0.582 \\
SegVol          & Box & 0.370 & 0.252 \\
VISTA3D         & Box & 0.190 & 0.119 \\
SAM-Med3D       & Box & 0.015 & 0.008 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Takeaway.}
On true nodules, healthy-counterfactual inpainting matches strong
2D interactive segmentation and outperforms several native-3D
promptable models. The next experiment asks whether that
competitiveness comes at the cost of over-calling on normal lung.

\subsection{Healthy False-Positive Probe}
Table~\ref{tab:healthy_fp} reports mean predicted mask volume
(mm\(^3\)) on random boxes in nodule-free lung. Among methods that
remain competitive on LIDC (Table~\ref{tab:main}), Ours yields the
\emph{smallest} spurious volume---substantially below SAM-Med2D,
SegVol, and especially nnInteractive. VISTA3D is nearly silent on
healthy tissue, but its LIDC Dice (\(0.19\)) shows that silence
alone is not useful without sensitivity to real nodules.

\begin{table}[t]
\centering
\caption{Healthy false-positive probe: random boxes in nodule-free
lung (\(N{=}3840\)). Lower predicted volume is better. Dice against
empty GT is omitted (near zero for any non-empty prediction).}
\label{tab:healthy_fp}
\begin{tabular}{lc}
\toprule
Method & Pred.\ volume (mm\(^3\)) \(\downarrow\) \\
\midrule
VISTA3D$^\dagger$     & \(0\) \\
\textbf{Ours (60-step, \(R{=}3\))} & \textbf{1006} \\
Ours (40-step, \(R{=}2\)) & 1076 \\
SegVol                & 1395 \\
SAM-Med2D             & 1980 \\
SAM-Med3D             & 2872 \\
nnInteractive         & 4015 \\
\bottomrule
\end{tabular}\\[0.25em]
{\footnotesize $^\dagger$Near-silent on healthy prompts, but poor on
true nodules (Table~\ref{tab:main}).}
\end{table}

\paragraph{Takeaway.}
Interactive segmenters optimize ``objectness in the box'' and
therefore over-segment healthy parenchyma. A healthy generative
prior is naturally more conservative: competitive Dice when an
anomaly is present, and markedly less spurious mass when it is
not---the central empirical claim of this paper.

\subsection{Ablation: Steps and Refine Rounds}
Table~\ref{tab:ablation} varies inference steps and refine rounds
\(R\) on the LIDC positive set (same checkpoint). Moderate compute
(\(40\) steps, \(R{=}2\)) is sufficient for near-peak Dice;
additional steps yield diminishing returns.

\begin{table}[t]
\centering
\caption{Ablation of inference steps and refine rounds on LIDC
positive test (\(N{=}1658\)).}
\label{tab:ablation}
\begin{tabular}{lcc}
\toprule
Configuration & Dice \(\uparrow\) & IoU \(\uparrow\) \\
\midrule
30-step (refine on) & 0.721 & 0.587 \\
40-step (refine on) & 0.726 & 0.593 \\
\textbf{40-step, \(R{=}2\)} & \textbf{0.734} & \textbf{0.600} \\
60-step (refine on) & 0.722 & 0.588 \\
80-step, \(R{=}2\) & 0.725 & 0.590 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Discussion}
Across Tables~\ref{tab:main}--\ref{tab:healthy_fp}, the operating
point we recommend (\(40\)-step, \(R{=}2\)) is nearly tied with
SAM-Med2D on positive Dice while producing roughly half the
spurious predicted volume of SAM-Med2D---and about one quarter
that of nnInteractive---on healthy random boxes. This supports
the problem reframing in Section~\ref{sec:intro}: for
region-prompted lung CAD, anomaly verification against a healthy
prior is a better inductive bias than interactive object
segmentation when false positives matter.

% =============================================================
\section{Scope, Limitations, and Conclusion}
\label{sec:conclusion}

\subsection{Scope}
Within lung CT, the framework is anomaly-agnostic: solid nodules,
part-solid nodules, GGOs, and any localized deviation from healthy
parenchyma are detected via the same residual mechanism. Cross-
organ and cross-modality scope is empirically demonstrated (C8,
C9). Extension to ultrasound and PET requires domain-specific
healthy-prior training and is reserved for future work. The
framework is applicable to any clinical dataset with bounding box
or centroid annotations and a healthy-tissue corpus for training.

\subsection{Limitations}
\begin{itemize}
    \item \textbf{Inference speed.} Iterative reverse diffusion
    with per-step mask updates and classifier guidance is slower
    than a single network forward pass. DDIM acceleration is
    applied but per-nodule latency exceeds nnU-Net/nnInteractive.
    Consistency-model distillation is planned future work.

    \item \textbf{Classifier dependency.} C4's classifier is
    trained on LIDC-IDRI, introducing a semi-supervised element;
    it must be retrained per new organ/modality for C8/C9.

    \item \textbf{Patch size.} The $64^3$ patch-level framework
    may fragment very large or diffuse lesions at boundaries;
    multi-scale hierarchical inference is future work.

    \item \textbf{Modality scope.} Cross-modality validation
    covers brain MRI only; ultrasound and PET are not evaluated
    in this paper.
\end{itemize}

\subsection{Conclusion}
We presented a unified framework for weakly supervised 3D anomaly
segmentation requiring no voxel-level annotations at any stage.
Our soft mask-coupled reverse diffusion (C1, C2), multi-timestep
discrepancy score and classifier guidance (C3, C4), synthetic
augmentation and hinge regularizer (C5, C6), and cross-dataset/-
organ/-modality generalisation (C7--C9) constitute a coherent
system coupled through the evolving soft mask $m_t$. Because the
model is trained only on healthy tissue and never on any specific
pathology, it is anomaly-agnostic by construction: the same
pipeline detects solid nodules, GGOs, liver lesions, and brain
tumours using only weak labels---without retraining on lesion
data. The framework outperforms interactive foundation models
specifically in the sub-6\,mm, low-contrast regime where
boundary-detection paradigms fail, and generalises across organs
and modalities by design.

% =============================================================
\section*{Acknowledgments}
[Omitted for anonymous review.]

% =============================================================
\bibliography{paper}

\end{document}
